canaper: Categorical analysis of neo- and paleo-endemism in R

### 2022-09-29 ms/ms.Qmd
Update MS, add SI

Running headline  
*canaper*

@@

3. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 15 min to reproduce a canonical study).

@@

Community data is provided as a `data.frame` or matrix, with species as columns and communities (also referred to as "sites" or "grid-cells") as rows.
In this case, the data must include both row names and column names.
Community data may also be input as a tibble, in which case site names must be indicated in a dedicated column (default column name `"site"`), rather than row names since tibbles lack row names.
Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
However, all calculations of PD and PE use only presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.

@@

The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with `canaper`, the test data from Phylocom [@Webb2008]:

@@

`cpr_iter_sim()` is not required for the CANAPE workflow, but is rather a diagnostic function for use with randomization algorithms that work by exchanging cells of the community matrix such as `swap` [@Gotelli2003] or `curveball` [@Strona2014].
Each such exchange is known as an "iteration".
It is generally impossible to know *a-priori* how many iterations are needed to completely randomize a given matrix; this number depends on properties of the matrix and the randomization algorithm.
Generally, larger matrices with more skew (overabundance of zeros) will require more iterations (ADD CITATION).

@@

The following code demonstrates usage of `cpr_rand_test()`, using the same example dataset as above:

@@

The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()` (note that for this small example, not all possible types of `endem_type` are produced):

@@

The following code demonstrates parallel computing:

@@

To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v`r R.Version()$major`.`r R.Version()$minor`.
Conducting CANAPE on Australian *Acacia* using `canaper` is simple, requiring only two commands:

@@

# 1. Run randomization test
acacia_rand_res <- cpr_rand_test(
  acacia$comm, acacia$phy,
  null_model = "curveball",
  n_reps = 999, n_iterations = 100000,
  tbl_out = TRUE
)

@@

`canaper` does not include any plotting functions to visualize the results.
Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr):

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors see cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(legend.position = "bottom", legend.title = element_blank())

@@

# Load CANAPE results from canaper and biodiverse
tar_load(c(
  acacia_canape_cpr,
  acacia_canape_biod,
  acacia_canape_comp),
  store = here::here("_targets")
)

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also vailable that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the  `cpr_make_pal()` function.

@@

When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells (@tbl-cpr-biod).
The total number of significant cells was nearly identical between the two (*n* = `r sig_tally$canaper$significant$n_sites` and `r sig_tally$biodiverse$significant$n_sites` sites, `canaper` and Biodiverse, respectively).
The `rand_structured` and `curveball` null models both preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.

@@

# Load times
tar_load(acacia_biod_elapsed_min)

# Enforce dependency on acacia_canape_cpr_1
tar_load(acacia_canape_cpr_1,
  store = here::here("_targets"))

@@

We are not aware of any other R packages that conduct the entire CANAPE pipeline automatically.
However, there is a large number of packages for analyzing species diversity [approximately 40 packages out of 15,300 as of 2019\; @Pavoine2020].
Some of those that are more closely related to `canaper` include the following.
`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
`vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
`vegan` includes by far the greatest number of algorithms for generating random communities (> 30) and is used by `canaper`.
`adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics but does not appear to be currently maintained and has been removed from CRAN.

@@

The `canaper` package enables CANAPE completely within R for the first time.
This will simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.

@@

All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
Source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
Usage of `canaper` is documented at https://docs.ropensci.org/canaper/.
The `acacia` dataset was originally published by @Mishler2014.

@@

tibble(
  Dataset = names(comms),
  `*n* sites` = map_dbl(comms, nrow),
  `*n* species (community matrix)` = map_dbl(comms, ncol),
  `*n* species (tree)` = map_dbl(phylos, ape::Ntip)
) %>%
  knitr::kable()
```

@@

ggsave(plot = fig, file = "ms/figure_1.pdf", width = 150, height = 150, units = "mm")

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors could use cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(legend.position = "bottom", legend.title = element_blank())

@@

ggsave(plot = fig, file = "ms/figure_2.pdf", width = 100, height = 200, units = "mm")

>>>>

Running headline: *canaper*

@@

3. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 10 min to reproduce a canonical study).

@@

Community data is provided as a `data.frame` or matrix, with species as columns and communities (also referred to as "sites" or "grid-cells") as rows.
In this case, the data must include both row names and column names.
Community data may also be input as a tibble, in which case site names must be indicated in a dedicated column (default column name `"site"`), rather than row names since tibbles lack row names.
Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
However, all calculations of PD and PE use only presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.
The `points2comm()` function of the `phyloregion` package may be used to convert raw occurrence data (e.g., latitude and longitude of species occurrences) to a community matrix [@Daru2020a].

@@

The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with `canaper`, the test data from Phylocom [@Webb2008].

@@

`cpr_iter_sim()` is not required for the CANAPE workflow, but is rather a diagnostic function for use with randomization algorithms that work by exchanging cells of the community matrix such as `swap` [@Gotelli2003] or `curveball` [@Strona2014].
Each such exchange is known as an "iteration".
It is generally impossible to know *a-priori* how many iterations are needed to completely randomize a given matrix; this number depends on properties of the matrix and the randomization algorithm.
Generally, larger matrices with more skew (overabundance of zeros) will require more iterations [@Miklos2004].

@@

The following code demonstrates usage of `cpr_rand_test()`, using the same example dataset as above.

@@

The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()` (note that for this small example, not all possible types of `endem_type` are produced).

@@

The following code demonstrates parallel computing.

@@

To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v`r R.Version()$major`.`r R.Version()$minor`.
We used the `curveball` randomization algorithm with 50,000 iterations, since `cpr_iter_sim()` indicated this was sufficient to randomize the matrix (see "Major Functions").
Once the randomization algorithm and its settings have been selected, CANAPE can be conducted with just two commands.

@@

# 1. Run randomization test
acacia_rand_res <- cpr_rand_test(
  acacia$comm, acacia$phy,
  null_model = "curveball",
  n_reps = 999, n_iterations = 50000,
  tbl_out = TRUE
)

@@

`canaper` does not include any plotting functions to visualize the results.
Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors see cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

@@

# Load CANAPE results from canaper and biodiverse
tar_load(c(
  acacia_canape_cpr,
  acacia_canape_biod,
  acacia_canape_comp,
  acacia_biod_elapsed_min,
  # Enforce dependency on acacia_canape_cpr_1 for calculating times
  acacia_canape_cpr_1),
  store = here::here("_targets")
)

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also vailable that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
The same plot visualized with CVD-safe colors is available in the online supporting material (Fig. S1).

@@

When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells (@tbl-cpr-biod).
The total number of significant cells was very similar (*n* = `r sig_tally$canaper$significant$n_sites` and `r sig_tally$biodiverse$significant$n_sites` sites, `canaper` and Biodiverse, respectively).
The `rand_structured` and `curveball` null models both preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.

@@

@@

We are not aware of any other R packages that conduct the entire CANAPE pipeline automatically.
However, there is a large number of packages for analyzing species diversity [approximately 40 packages out of 15,300 as of 2019\; @Pavoine2020].
Some of those that are more closely related to `canaper` include the following.
`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
`vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
`vegan` includes by far the greatest number of algorithms for generating random communities (> 30) and is used by `canaper`.
`adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics.
`EcoSimR` [@Gotelli2013a] provides randomization algorithms as well as functions for characterizing community matrices and null distributions.
`PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.

@@

The `canaper` package enables CANAPE completely within R for the first time.
This will simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
`canaper` has already been used at least three studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a].
We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.

@@

All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
Source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
The `acacia` dataset was originally published by @Mishler2014.

@@

tibble(
  Dataset = names(comms) %>% paste0("`", ., "`"),
  `*n* sites` = map_dbl(comms, nrow),
  `*n* species (community matrix)` = map_dbl(comms, ncol),
  `*n* species (phylogeny)` = map_dbl(phylos, ape::Ntip)
) %>%
  knitr::kable()
```

@@

ggsave(plot = fig, file = here::here("ms/figure_1.pdf"),
  width = 150, height = 150, units = "mm")

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors could use cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

@@

ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
  width = 100, height = 200, units = "mm")
